Guarded Recursion in Agda via Sized Types

### 2019-01-24 code-oneclock-paper/latex/Introduction.tex
finished intro

Possibly non-terminating computations
and infinite structures such as non-wellfounded trees can be
represented in type theory by extending the type system with
coinductive types. Elements of these types, which we call processes,
are required to be productive \cite{Coquand93} in the sense that every
finite unfolding can be computed in a finite number of steps.
In Agda's encoding of coinductive types using ``musical notation''
\cite{DA10} \NV{What about Coq?}, productivity is achieved using a strict method: in the
definition of corecursive functions, the recursive
call is required to appear directly under
the application of a constructor. This restriction
typically makes programming with processes cumbersome,
which spawned the search for alternative techniques for ensuring the
well-definedness of corecursive definitions.

@@

Using techniques for representing ``type theory in type theory''
\cite{Chapman09,AltenkirchK16}, in Section \ref{sec:syntax} we present
an Agda formalization of the syntax of a simple type theory with
guarded recursion. This object language, that we call \GTT, is a variant of Atkey and
McBride's type system for productive coprogramming
\cite{atkey2013productive} in which the clock context can contain at
most one clock.

In Section \ref{sec:presheaf_sem}, we develop a denotational semantics
for \GTT\ using sized types. More precisely, we define a presheaf
model where the type formers of simply typed lambda calculus are
evaluated using the standard Kripke semantics. Typically the semantics
of type theories for guarded recursion is given in the topos of
trees or variations thereof \cite{BMSS-synthetic} \NV{Add citations here}.
In order to make clear the connections between guarded recursion and
sized types, we take the preorder of sizes as the indexing
category of our presheaves. This means that types and contexts of \GTT\ are
interpreted as antitone sized types.

\NV{WIP from this point}

@@

In contrast, interpreting guarded recursion and guarded recursive types is more challenging.
The most common model in the literature is the topos of trees \cite{BMSS-synthetic}.
This is a presheaf topos, whose objects are sets indexed by a natural number.
We use a slightly different approach: instead of natural numbers, we use sizes.
This means that the required interpretations need to be modified.

Our contributions are as follows
\begin{enumerate}
	\item We give a new denotational semantics of guarded recursive type theory in type theory;
	\item We show that guarded recursion can be interpreted in sized types.
	\item We formally prove the syntax is consistent.
\end{enumerate}

In Section 2, we describe the meta-theory in which we work.
In Section 3, we describe the syntax of the object theory, which is basically a variant of guarded recursion as done by Atkey and McBride \cite{atkey2013productive}.
We do this in the form of a deep embedding.
More specifically, we define data types representing the contexts, types, terms, and so on, and we give the rules for definitional equality.
In Sections 4 and 5, we define the denotational semantics of this syntax.
First, we give the interpretation of context, types, and terms in the model, and we show how to interpret simple type formers.
Then we show this model supports guarded recursion and guarded recursive types.
Finally, we show the syntax is consistent in Section 6.

>>>>

Possibly non-terminating computations
and infinite structures such as non-wellfounded trees can be
represented in type theory by extending the type system with
coinductive types. Recursively defined elements of these types
are required to be productive \cite{Coquand93} in the sense that every
finite unfolding can be computed in a finite number of steps.
In Agda's encoding of coinductive types using ``musical notation''
\cite{DA10}, productivity is enforced via a strict obligation: in the
definition of a corecursive function, the recursive
call is required to appear directly under
the application of a constructor. A similar syntactic
check is employed in Coq.
This restriction
typically makes programming with coinductive types cumbersome,
which spawned the search for alternative techniques for ensuring the
well-definedness of corecursive definitions.

@@

Utilizing techniques for representing ``type theory in type theory''
\cite{Chapman09,AltenkirchK16}, we present
an Agda formalization of the syntax of a simple type theory with
guarded recursion. This object language, that we call \GTT, is a variant of Atkey and
McBride's type system for productive coprogramming
\cite{atkey2013productive} in which the clock context can contain at
most one clock. The types of \GTT\ include the aformentioned \IC{▻} modality and a
\IC{□}  modality, the nameless analogue of  Atkey and
McBride's universal clock quantification.

Afterwards we develop a categorical semantics
for \GTT\ using sized types. More precisely, we define a presheaf
model where the type formers of simply typed lambda calculus are
evaluated using the standard Kripke semantics. Typically the semantics
of type theories for guarded recursion is given in the topos of
trees or variations thereof \cite{BMSS-synthetic,Mogelberg14,MannaaM18}.
Here, in order to make clear the connections between guarded recursion and
sized types, we take the preorder of sizes as the indexing
category of our presheaves. This means that types and contexts of \GTT\ are
interpreted as antitone sized types. 
The well-order relation on sizes is fundamental for giving a
terminating definition of the semantic fixpoint combinator.

%% A detailed account of how to model the characteristic features of
%% \GTT, consisting of the \IC{▻} and
%% \IC{□}  modality together with guarded recursive types, is given in
%% Section \ref{sec:guarded}.

To summarize, the contributions of this paper are twofold:
\begin{enumerate}
  \item We formalize syntax and semantics of a type theory with
    guarded recursion in Agda. This is the first such denotational model developed using an extension of Martin-L\"of intensional
    type theory as the metalanguage, in contrast with the previous
    set-theoretic models of guarded recursion.
\item The interpretation of the characteristic features of guarded
  recursion crucially requires the employment of sized types. This shows that
  guarded recursive types can be reduced to sized types and
  constitutes a stepping stone towards a clear understanding of  the connections
  between different approaches to productive coprogramming.
%	\item We formally prove the syntax is consistent.
\end{enumerate}

This paper only include the essential parts of our Agda
formalization. The full code is available at
\url{https://github.com/niccoloveltri/agda-clott}. The formalization
uses Agda 2.5.4.1 and Agda standard library 0.16.

%\NV{WIP from this point}

@@

%% In contrast, interpreting guarded recursion and guarded recursive types is more challenging.
%% The most common model in the literature is the topos of trees \cite{BMSS-synthetic}.
%% This is a presheaf topos, whose objects are sets indexed by a natural number.
%% We use a slightly different approach: instead of natural numbers, we use sizes.
%% This means that the required interpretations need to be modified.

%% Our contributions are as follows

%% In Section 2, we describe the meta-theory in which we work.
%% In Section 3, we describe the syntax of the object theory, which is basically a variant of guarded recursion as done by Atkey and McBride \cite{atkey2013productive}.
%% We do this in the form of a deep embedding.
%% More specifically, we define data types representing the contexts, types, terms, and so on, and we give the rules for definitional equality.
%% In Sections 4 and 5, we define the denotational semantics of this syntax.
%% First, we give the interpretation of context, types, and terms in the model, and we show how to interpret simple type formers.
%% Then we show this model supports guarded recursion and guarded recursive types.
%% Finally, we show the syntax is consistent in Section 6.

### 2019-01-24 code-oneclock-paper/latex/paper.tex
rephrasing intro
WIP: intro
finished intro

\acknowledgements{We are thankful to Andreas Abel, Herman Geuvers,
  Rasmus Ejlers M{\o}gelberg and Andrea Vezzosi for discussions and
valuable hints.}%optional

@@

\newunicodechar{⊤}{$\top$}
\newunicodechar{≡}{$\equiv$}
\newunicodechar{₁}{$_1$}
\newunicodechar{₂}{$_2$}
\newunicodechar{₃}{$_3$}
\newunicodechar{Γ}{$\Gamma$}
\newunicodechar{Δ}{$\Delta$}
\newunicodechar{Σ}{$\Sigma$}
\newunicodechar{α}{$\alpha$}
\newunicodechar{β}{$\beta$}
\newunicodechar{η}{$\eta$}
\newunicodechar{γ}{$\gamma$}
\newunicodechar{ε}{$\varepsilon$}
\newunicodechar{θ}{$\theta$}
\newunicodechar{λ}{$\lambda$}
\newunicodechar{μ}{$\mu$}
\newunicodechar{κ}{$\kappa$}
\newunicodechar{π}{$\pi$}
\newunicodechar{∈}{$\in$}
\newunicodechar{⊞}{$\boxplus$}
\newunicodechar{⊠}{$\boxtimes$}
\newunicodechar{∁}{$\mathbb{C}$}
\newunicodechar{►}{$\blacktriangleright$}
\newunicodechar{▻}{$\vartriangleright$}
\newunicodechar{𝟙}{\textbf{1}}
\newunicodechar{∙}{$\cdot$}
\newunicodechar{∀}{$\forall$}
\newunicodechar{ℓ}{$\ell$}
\newunicodechar{∅}{$\varnothing$}
\newunicodechar{⟶}{$\rightarrow$}
\newunicodechar{□}{$\square$}
\newunicodechar{⇡}{$\uparrow$}
\newunicodechar{⊛}{$\circledast$}
\newunicodechar{⊎}{$\uplus$}
\newunicodechar{⟨}{$\langle$}
\newunicodechar{⟩}{$\rangle$}
\newunicodechar{■}{$\blacksquare$}
\newunicodechar{∎}{$\blacksquare$}
\newunicodechar{⇒}{$\Rightarrow$}
\newunicodechar{⊕}{$\oplus$}
\newunicodechar{⊗}{$\otimes$}
\newunicodechar{∼}{$\sim$}
\newunicodechar{≈}{$\approx$}
\newunicodechar{⊔}{$\sqcup$}
\newunicodechar{⟦}{$\llbracket$}
\newunicodechar{⟧}{$\rrbracket$}
\newunicodechar{⊥}{$\bot$}
\newunicodechar{∞}{$\infty$}
\newunicodechar{⇑}{$\Uparrow$}
% Add more as you need them (shouldn't happen often).

@@

\section{Introduction}
\label{sec:intro}
\input{Introduction}
\nocite{*}

@@

\section{Deep Embedding}
\label{sec:syntax}
\input{Prelude/Syntax}
\input{Streams}

\section{Categorical Semantics}
Next we define categorical semantics for the syntax defined in \Cref{sec:syntax}.
To do so, we use a similar method as used by M{\o}gelberg \cite{Mogelberg14}.
The interpretation of types depend on the clock contexts.
If the clock context is empty, then types become sets.
If the clock context has one clock, then types are interpreted as objects in the topos of trees.

@@

\subsection{Simple Type Theory in Presheaves}
%\label{sec:kripke}
\input{CloTT/Structure/Contexts}

>>>>

\acknowledgements{We are thankful to Andreas Abel, Guillaume Allais, Herman Geuvers,
  Rasmus Ejlers M{\o}gelberg and Andrea Vezzosi for discussions and
valuable hints.}%optional

@@

\newunicodechar{⊤}{$\top$}
\newunicodechar{≡}{$\equiv$}
\newunicodechar{₁}{$_1$}
\newunicodechar{₂}{$_2$}
\newunicodechar{₃}{$_3$}
\newunicodechar{Γ}{$\Gamma$}
\newunicodechar{Δ}{$\Delta$}
\newunicodechar{Σ}{$\Sigma$}
\newunicodechar{α}{$\alpha$}
\newunicodechar{β}{$\beta$}
\newunicodechar{η}{$\eta$}
\newunicodechar{γ}{$\gamma$}
\newunicodechar{ε}{$\varepsilon$}
\newunicodechar{θ}{$\theta$}
\newunicodechar{λ}{$\lambda$}
\newunicodechar{μ}{$\mu$}
\newunicodechar{κ}{$\kappa$}
\newunicodechar{π}{$\pi$}
\newunicodechar{∈}{$\in$}
\newunicodechar{⊞}{$\boxplus$}
\newunicodechar{⊠}{$\boxtimes$}
\newunicodechar{∁}{$\mathbb{C}$}
\newunicodechar{►}{$\blacktriangleright$}
\newunicodechar{▻}{$\vartriangleright$}
\newunicodechar{𝟙}{\textbf{1}}
\newunicodechar{∙}{$\cdot$}
\newunicodechar{∀}{$\forall$}
\newunicodechar{ℓ}{$\ell$}
\newunicodechar{∅}{$\varnothing$}
\newunicodechar{⟶}{$\rightarrow$}
\newunicodechar{□}{$\square$}
\newunicodechar{⇡}{$\uparrow$}
\newunicodechar{⊛}{$\circledast$}
\newunicodechar{⊎}{$\uplus$}
\newunicodechar{⟨}{$\langle$}
\newunicodechar{⟩}{$\rangle$}
\newunicodechar{■}{$\blacksquare$}
\newunicodechar{∎}{$\blacksquare$}
\newunicodechar{⇒}{$\Rightarrow$}
\newunicodechar{⊕}{$\oplus$}
\newunicodechar{⊗}{$\otimes$}
\newunicodechar{∼}{$\sim$}
\newunicodechar{≈}{$\approx$}
\newunicodechar{⊔}{$\sqcup$}
\newunicodechar{⟦}{$\llbracket$}
\newunicodechar{⟧}{$\rrbracket$}
\newunicodechar{⊥}{$\bot$}
\newunicodechar{∞}{$\infty$}
\newunicodechar{⇑}{$\Uparrow$}
\newunicodechar{≃}{$\simeq$}
% Add more as you need them (shouldn't happen often).

@@

\section{Introduction}
\label{sec:intro}
\input{Introduction}
%\nocite{*}

@@

\section{The Object Type Theory}
\label{sec:syntax}
\input{Prelude/Syntax}
\input{Streams}

\section{Categorical Semantics}

Next we define categorical semantics for the syntax defined in \Cref{sec:syntax}.
To do so, we use a similar method as used by M{\o}gelberg \cite{Mogelberg14}.
The interpretation of types depend on the clock contexts.
If the clock context is empty, then types become sets.
If the clock context has one clock, then types are interpreted as objects in the topos of trees.

@@

\subsection{Modelling Simple Type Theory}
%\label{sec:kripke}
\input{CloTT/Structure/Contexts}
